### 2016-08-03 latex/input/abstract.tex
added script for generating plots of GICF runs; updated paper

We demonstrate a computational method to predict the clinical
phenotypes of a patient from raw metagenomic sequence read data. We  compared 
two state of the art programs for annotating the sequence data, 
UCLUST and Kraken, and using their output for feature generation. 
We apply these programs 
to a set of over 1.3 million reads from 
904 patients, some of whom have liver cirrhosis, encephalopathy due to liver cirrhosis, or 
neither disease. Once the reads have been processed by UCLUST or 
Kraken, we use  Support Vector Machines to setup the 
clinical phenotype prediction problem.  
%
We find that too many false negatives are being predicted by 
the classifier. In order to address the issue, we scale features to
improve the classification model and evaluate the 
end results on a held-out test set. We
demonstrate our approach works quickly and 
accurately with an 85.64\% success rate when we use the 
UCLUST representation. We also find that UCLUST generally performs better than Kraken, with the latter having an 80.66\% success rate. We also test our classifier on several subsets of the data, with success rates ranging from 69.81\% to 96.72\%.

>>>>

The recent advent of Metagenome-Wide Association Studies (MGWAS) has allowed for increased accuracy in the prediction of patient phenotype (disease), but has also presented big data challenges. Meanwhile, Multiple Instance Learning (MIL) is useful in the domain of bioinformatics because, in addition to classifying patient phenotype, it can also identify individual parts of the metagenome that are indicative of that phenotype, leading to better understanding of the disease. We demonstrate a novel, efficient, and effective MIL-based computational pipeline to predict patient phenotype from MGWAS data. Specifically, we use a Distance-based Bag of Words method, which has been shown to be one of the most effective and efficient MIL methods. This involves assembly of the metagenomic data, clustering of the assembled contigs, extracting features from the contigs, and using a standard SVM classifier to predict patient labels and identify the most relevant read clusters. With the exception of the given labels for the patients, this entire process is \emph{de novo}. We use data from a well-known MGWAS study of patients with Type-2 Diabetes and show that our pipeline significantly outperforms the classifier used in that paper.

### 2016-08-03 latex/input/conclusion.tex
added script for generating plots of GICF runs; updated paper

We have demonstrated a computational method 
to predict clinical phenotypes using metagenomic sequence data with intermediate 
OTU representation using state-of-the-art binning and taxonomic profiling approaches. 
%
%
Our results showed that 
the UCLUST representation 
achieved 85.64\% accuracy and 70.46\% F1-Score. Kraken performed less well, with 80.66\% accuracy and 53.34\% F1-Score. 
For the 1.3 million sequence reads UCLUST ran in 180 seconds and Kraken for 120 seconds, indicating that both performed efficiently enough to feasibly use them with larger data sets.

We then used UCLUST to test three one-versus-one clinical phenotype classifiers,
which generally produced stronger results than our original 
comparison of the "Encephalopathy" class versus all other 
classes. In particular, the classification of 
"Encephalopathy" versus "Control" performed the best. This intuitively seems reasonable. Whereas 
the original comparison requires making distinctions between 
those who have both encephalopathy and liver cirrhosis, those who have only the latter, and those who have neither, this pairwise comparison focuses on the simpler distinction 
between patients who have both diseases and those who have neither. The other two pairwise comparisons also similarly simplify the classification problem. When Kraken was used for the three pairwise comparisons, it had the peculiar effect of drastically increasing the recall while maintaining roughly the same accuracy and precision. This is most likely due to the fact that the pairwise comparisons involved data sets that were less negatively-skewed and thus didn't cause the classifier to predict as negatively based on Kraken's results. This is combined with the relatively strong accuracy and precision numbers in Kraken's original run, versus its poor recall numbers that left plenty of room for improvement.

Our technique of applying a multiplier to feature values for positive examples produced the effect of improving the recall in most cases. These results were generally corroborated by similar improvements in the cross validation results. Another effect was that precision and accuracy results were also often improved after applying this technique. This intuitively makes sense, as switching false negatives to true positives also positively affects the accuracy and precision ratios. Overall, this feature engineering process appeared to have a generally positive effect on most of the results, suggesting that similar techniques for improving classification results merit further investigation.

Based on our results demonstrating over 85\% accuracy for the original run and even better results for some of the pairwise comparisons, we believe the use of machine learning
classification techniques to predict clinical phenotypes 
merits continued research. There are many potential practical 
uses of such technology for medical purposes, including 
diagnostics and research into diseases and their 
relationship with the human microbiome.

\subsection{Future Work}

Kraken is developed to be fast, accurate, and precise, all 
of which we found to be true. However, it was not as accurate as a representation in comparison to 
UCLUST. Kraken's poor performance was mostly due to  poor matching of $k$-mers in the reads to taxonomic classes in its database. Often, a read was only a one or two percent match to the class that Kraken labeled it as. This 
may be due in part to database constraints. Kraken's full database is very computationally expensive to download and build. We 
used a 4 GB custom subset of the full 160 GB database. It is 
possible that the full database would have led to improved performance, and this is 
a future research question.

There is also an opportunity to use a much larger set of sequence reads. Having more training examples for the classifier would  lead to a stronger and more generalizable model for predicting clinical phenotypes. The use of other annotation and classification methods could also be explored. Finally, a limited set of feature selection and scaling techniques were explored, and future research could investigate further how such techniques could be used to improve predictions.

>>>>

We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data. We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study. Finally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods. Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level labels and validate those labels against clinically verified labels.

### 2016-08-03 latex/input/experiments.tex
added script for generating plots of GICF runs; updated paper

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{./diagrams/PreFeatEng}
\caption{This diagram shows the results for the comparison between UCLUST and Kraken when classifying "Encephalopathy" versus all other classes. UCLUST had an accuracy and F1-Score of 84.53\% and 67.44\% respectively. Kraken didn't perform as well, with an accuracy of 79.56\% and an F1-Score of 43.08\%. \label{prefeature}}
\end{figure}

\begin{table*}[t]
\begin{center}
\caption{Comparative Run Time of UCLUST and Kraken Representations. \label{tab:time}}
\begin{tabular}{|c|c|}\hline
Representation & Time\\\hline
UCLUST & 180s\\\hline
Kraken & 120s\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Comparative Performance of UCLUST and Kraken Representations. 
\label{tab:comp}}
\begin{tabular}{|c|cccc|cccc|}\hline
Representation & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
UCLUST & 84.53 & 78.38 & 59.18 & 67.44 & 88.26 & 83.22 & 67.39 & 74.47\\\hline
Kraken & 79.56 & 87.5 & 28.57 & 43.08 & 78.18 & 70.31 & 24.46 & 36.29\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Three Pairwise Comparisons Using UCLUST. \label{tab:pairuclust}}
\begin{tabular}{|c|cccc|cccc|}\hline
Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
Encephalopathy vs. No Encephalopathy & 88.68 & 90 & 86.54 & 88.23 & 84.47 & 81.77 & 81.77 & 81.77\\\hline
Encephalopathy vs. Control & 96.72 & 95.35 & 100 & 97.62 & 94.26 & 93.2 & 100 & 96.48\\\hline
No Encephalopathy vs. Control & 83.78 & 82.61 & 100 & 90.48 & 87.5 & 86.96 & 99.59 & 92.85\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Three Pairwise Comparisons Using Kraken. \label{tab:pairkraken}}
\begin{tabular}{|c|cccc|cccc|}\hline
Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
Encephalopathy vs. No Encephalopathy & 67.92 & 82.14 & 44.23 & 57.50 & 68.94 & 72.07 & 44.20 & 54.79\\\hline
Encephalopathy vs. Control & 78.69 & 76.92 & 97.56 & 86.02 & 79.51 & 81.14 & 96.35 & 88.09\\\hline
No Encephalopathy vs. Control & 77.03 & 77.03 & 100 & 87.02 & 81.42 & 81.42 & 100 & 89.76\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Comparative Performance After Feature Engineering. \label{tab:pfe}}
\begin{tabular}{|c|cccc|cccc|}\hline
Representation & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
UCLUST & 85.64 & 79.49 & 63.27 & 70.46 & 88.54 & 82.58 & 69.57 & 75.52\\\hline
Kraken & 80.66 & 76.92 & 40.82 & 53.34 & 80.8 & 75.86 & 35.87 & 48.93\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Three Pairwise Comparisons Using UCLUST After Feature Engineering. \label{tab:pairuclustfe}}
\begin{tabular}{|c|cccc|cccc|}\hline
Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
Encephalopathy vs. No Encephalopathy & 88.68 & 90 & 86.54 & 88.23 & 86.35 & 83.24 & 85.08 & 84.15\\\hline
Encephalopathy vs. Control & 96.72 & 95.35 & 100 & 97.62 & 95.08 & 94.12 & 100 & 96.97\\\hline
No Encephalopathy vs. Control & 87.84 & 86.36 & 100 & 92.68 & 91.22 & 90.57 & 99.59 & 95.05\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Three Pairwise Comparisons Using Kraken After Feature Engineering. \label{tab:pairkrakenfe}}
\begin{tabular}{|c|cccc|cccc|}\hline
Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
Encephalopathy vs. No Encephalopathy & 69.81 & 83.33 & 48.08 & 60.98 & 77.18 & 80.88 & 60.77 & 69.40\\\hline
Encephalopathy vs. Control & 86.89 & 86.67 & 95.12 & 90.70 & 87.30 & 88.15 & 96.88 & 92.31\\\hline
No Encephalopathy vs. Control & 77.03 & 77.78 & 98.25 & 86.82 & 82.77 & 82.53 & 100 & 90.43\\\hline
\end{tabular}
\end{center}
\end{table*}

Table \ref{tab:time} shows 
the run time of each representation in seconds. We had total of 1.3 million sequence reads. UCLUST unsupervised clustering 
took 180 seconds, while Kraken's OTU database lookups took 120 seconds.

Figure \ref{prefeature} and Table \ref{tab:comp} show 
the classification performance of UCLUST and 
Kraken in distinguishing patients in the "Encephalopathy" class versus 
the other classes. UCLUST correctly predicted the clinical 
phenotype 84.53\% of the time, and had a precision of 78.38\% and recall of 59.18\% on the held-out set. LOOCV results were 
slightly better for each metric than results on the test 
set by about 4-8\%. Kraken had an accuracy of 79.56\% with an even 
larger disparity between precision and recall (87.5\% for the former and 
28.57\% for the latter), which led to a relatively poor F1-score of 43.08\%. LOOCV results for Kraken 
were worse by 1.38\% for accuracy, 17.19\% for precision, 4.11\% for recall, and 6.79\% for F1-Score. As 
the results show, UCLUST generally performed better as a feature 
representation, with 4.97\% higher accuracy and 24.36\% higher F1-score.

Table \ref{tab:pairuclust} displays the performance of UCLUST-based classification 
with regards to training three one-versus-one clinical phenotypic classifiers.  The first comparison 
distinguished between the "Encephalopathy" class and the "No Encephalopathy" class, effectively determining whether or not a patient who had liver cirrhosis also had encephalopathy. Results were much improved when compared with the findings in table \ref{tab:comp}. The accuracy, precision, recall, and F1-score were all  86-90\%. LOOCV
results were worse by about 4-8\% per metric. The 
second comparison distinguished between the "Encephalopathy" class and 
the "Control" class, effectively determining whether a 
patient had encephalopathy caused by liver cirrhosis or had neither disease. This comparison had the best results of the ones we examined, with greater than 95\% performance in all metrics and 100\% recall. The cross validation
results were about 1-2\% worse for each metric except recall, which was still 100\%. The 
third comparison distinguished between the "No Encephalopathy and liver cirrhosis" class and the "Control" class, effectively determining which among the patients lacking encephalopathy had liver cirrhosis. This comparison had an accuracy of 83.78\% and precision of 82.61\%, but had a 100\% recall which led to a high F1-score of 90.48\%. Cross validation results were better by 3.72\% for accuracy, 4.35\% for precision, and 2.37\% better for F1-Score, with essentially the same recall.  Apart from the accuracy and precision of the third comparison, these three pairwise comparisons generally performed better across the board than the results in Table \ref{tab:comp}, with the second pairwise comparison having the best performance.

Table \ref{tab:pairkraken} displays the performance of Kraken with regards to the same three pairwise comparisons described above. For the Encephalopathy vs. No Encephalopathy run, the accuracy was 67.92\%, with 82.14\% precision, 44.23\% recall, and 57.50\% F1-Score. Cross validation produced a precision that was 10.07\% lower but largely the same accuracy and recall. The Encephalopathy vs. Control run produced good results, with 78.69\% accuracy, 76.92\% precision, 97.56\% recall, and 86.02\% F1-Score. LOOCV results mostly confirmed those results, except the precision, which was 4.22\% higher. The No Encephalopathy vs. Control run had quite similar results to the Encephalopathy vs. Control run. The main pattern that appears to emerge from these results is that the recall for Kraken was generally far superior in the pairwise runs than in the original run, which increased the F1-Scores but resulted in similar accuracy and precision. The Encephalopathy vs. No Encephalopathy run was significantly worse than the other two.

Table \ref{tab:pfe} shows the results for the same setup as 
in Table \ref{tab:comp} after we performed feature engineering. For both representations in Table \ref{tab:comp}, the recall was significantly lower than the precision, indicating that there were many more false negatives than false positives. We thus engineered the features in the training set. Certain features in the training set were most strongly present in positive patients, and as such were correlated with positive examples. Some examples in the test set had slightly lower values for those features, and were thus labelled by the classifier as negative examples, when in reality they were positive. We aimed to influence some of these borderline false negative predictions by lowering the threshold for those positively correlated features that would lead to an example being predicted as positive. In order to reduce the threshold for those positively correlated features, we applied a multiplier of 0.9 to the values of each feature of every positive patient in the training set. This had the effect of causing some previously borderline negative predictions to become positive predictions, primarily reducing the number of false negatives and improving the recall scores, but also positively impacting most other metrics as well.  Attempts to use a multiplier of 
less than 0.9 did not appear to further 
improve the results.

With the multiplier of 0.9 applied to the 
feature values for positive training examples, the recall 
increased by 4.09\% for UCLUST and by 12.25\% for Kraken. This was confirmed in the 
cross validation results as well, 
in which UCLUST's recall increased by 2.18\% and 
Kraken's recall increased by 11.41\%. Because of the improvement 
in recall, both representations also had similarly improved F1-Scores after feature engineering. Accuracy and precision remained generally consistent. 
%
Kraken's LOOCV precision 
increased by 5.55\%. This suggests that Kraken's very high precision was tied to its lack of many positive phenotype predictions in general, resulting in a very low number of false positives. While Kraken improved more than UCLUST did due to feature engineering, UCLUST still performed better in all metrics.

Tables \ref{tab:pairuclustfe} and \ref{tab:pairkrakenfe} display the results after the same feature engineering techniques were applied to the pairwise results in \ref{tab:pairuclust} and \ref{tab:pairkraken}, respectively. UCLUST's No Encephalopathy vs. Control run saw an increase of 4.06\% in accuracy, 3.75\% in precision, and similar gains in its cross validation results for accuracy and precision. UCLUST's other two runs were largely unaffected, albeit with slightly better cross validation results. Kraken's Encephalopathy vs. Control run was greatly improved after feature engineering, with an 8.20\% increase in accuracy and a 9.75\% increase in precision, with similar increases in cross validation results. This more than offset a slight decrease in recall. The Encephalopathy vs. No Encephalopathy run had mostly unchanged results, albeit with much improved cross validation scores, and the No Encephalopathy vs. Control run was not significantly affected.

>>>>

\subsection{Data, Software, and Hardware Used}

Our main dataset comes from a GWAS of Type 2 Diabetes (T2D) in Chinese adults, published by Qin et al \cite{qin041012}. The data is publicly available on the National Center for Biotechnology Information (NCBI) \cite{ncbi} website using the accession number SRP011011. The full dataset contains over 200 patients, but for our study we used a subset of 60 patients due to lengthy download and format conversion times. Of these 60 patients, 30 had T2D and 30 were controls. For the purpose of testing our cluster sampling methods, we also used a smaller dataset obtained from Dr. Patrick Gillevet of GMU, which is not publicly available. It contains 904 patients, 239 of whom have encephalopathy and 665 of whom are controls, but is much smaller because it uses 16S rRNA instead of whole genome data. The encephalopathy dataset contains about 1.3 million reads, while the T2D dataset contains about 1.3 billion reads.

As previously mentioned, we use the publicly-available UCLUST \cite{Edgar10} software for clustering. For classification, we use Support Vector Machines (SVMs) \cite{vap95}, one of the most powerful and versatile binary classifiers used 
in myriad applications. 
%
For a  binary classification problem, the SVM framework 
maximizes the separation between the two classes. In addition to the ubiquity and effectiveness of SVMs, they are well established in both supervised and multiple instance learning, making it easy to compare the two approaches. For our standard supervised SVM, we use svm-light \cite{joachims08}, a popular and publicly available implementation. For multiple instance learning, we use the python implementation by G. Doran and S. Ray ("MISVM") \cite{doran122013} for its ease of use and public availability.

\subsection{Efficient Unsupervised Clustering using UCLUST}
\label{uclustmethods}

From early experimental results, UCLUST can still be computationally infeasible for very large datasets such as the one we will be using, despite its generally excellent speed. For our T2D dataset, UCLUST was only able to cluster 2.8\% of the reads within 12 hours, at which point ARGO killed the batch job. UCLUST slows down as more clusters are formed, as each new read needs to be compared to more existing clusters. Thus, as the number of clusters grows, the execution time of UCLUST increases rapidly. The only way to feasibly complete the clustering phase in a reasonable amount of time was to cluster the patients individually in parallel, instead of clustering the reads of all patients together. However, this leads to the problem of having mismatched clusterings between patients, that is, cluster X from patient A may represent the same OTU as cluster Y from patient B. The solution is to compare the clusters from different patients and map them to each other, and then to combine clusters from different patients that represent the same OTU and output a single clustering output file for all patients. However, comparing every read from each cluster to every read from all other clusters is also computationally infeasible. Thus, we developed methods to sample representative reads from each patient and use them to create the consensus clustering. We explored a number of different methods, but two of them in particular were able to achieve significant speed improvements without significantly harming classification results. For each of them, we used a parameter "alpha", which was used to select 1 out of every alpha reads. For instance, if alpha=10, 1/10 of reads would be used.

We refer to one of our methods as "seed re-clustering" (UCLUST refers to cluster centroids as "seeds"). In this method, the reads corresponding to the cluster seeds from each patient are copied into a temporary file, and then those reads are clustered. This runs quickly, since clusters in this setting tend to have hundreds to thousands of reads, so the number of cluster seeds is relatively small. We then construct a map, which maps each seed from its original cluster number to its new cluster number from the seed clustering. We go back to the individual patient clustering output, and select 1 out of every alpha lines to be placed in the consensus clustering file. As we write the lines to the consensus file, we use the map we developed to change the cluster number for the line from the original clustering number to the new consensus cluster number. We refer to the other method as "two round clustering". We select one out of every alpha reads from the clustering output of each patient, place them in a single file, and run UCLUST on this smaller subset to arrive at a consensus clustering. The main difference between two round clustering and seed re-clustering is that here we select a distribution of reads from the original clustering to determine the consensus clustering, instead of cluster seeds/centroids.

We sought to empirically evaluate how these methods affected clustering runtime and classification accuracy, since feature vectors are generated from the clustering output and thus classification results are affected by the clustering methods. Without use of these sampling methods, UCLUST didn't finish within ARGO's 12 hour time limit on the T2D dataset, so we couldn't evaluate the classification results on that dataset and compare them to the classification results with the clustering sample methods. In order to evaluate these methods, we used the aforementioned Encephalopathy dataset, which is much smaller. We measured the time it took to perform UCLUST on all patients at once, as well as the time taken by each of our two sampling methods. We then used the respective clustering outputs to generate feature vectors and run svm-light classification (as described in the next section) and compared the results, which are discussed in the "Experimental Results" section. For UCLUST, default settings were used, except the threshold for placing two sequences in the same cluster was set to 40\% match (--id 0.40), and the --usersort option was used.

\subsection{Supervised and Multiple Instance Classification using Support Vector Machines (SVMs)}
\label{svm}

For our formulation we use SVMs to make clinical phenotype prediction for each patient sample based on their  input metagenome sequence samples represented with their clustering or taxa membership features. With our clustering approach,  we have a set of reads that have been assigned to clusters based on UCLUST and our techniques. We now generate feature vectors from each patient from this clustering output. We follow a different process of feature vector generation for svm-light and MISVM. For each method, we read through the consensus clustering output from UCLUST line-by-line, forming feature vectors as we go. UCLUST output lines provide the cluster number, the percentage match to the cluster seed, and the patient that read comes from.

For svm-light, we have one feature vector per patient. We use each cluster/OTU as a feature, meaning that the presence of various microbes are the features for the patient. Each number (feature) in the vector corresponds to a cluster/OTU, and the value for that number corresponds to how closely the read from the patient matched the cluster seed. For each input 
sequence, we use the percentage match to its cluster seed as 
the value for that feature and average the 
sequence match if multiple reads from the metagenome samples 
are assigned to the same cluster by UCLUST. Each sequence read includes a label for the patient that it came from, so each read only affects the feature vector corresponding to the patient that it came from. For example, say patient A has a read with 95\% match to OTU 2, a read that is the cluster seed of OTU 3, a read that is a 93\% match to OTU 2, a read that is a 97\% match to OTU 5, and no reads for OTU 1 or 4. Patient A's feature vector would then look like: [0 0.94 1 0 0.97]. We first initialize the patient feature vectors to all 0s and then iterate through the UCLUST output lines and update the feature vectors as we go. For MISVM, patients are bags, reads for patients are instances, and each read/instance has a feature vector. We iterate through the UCLUST output. For each read, we add that read as an instance to the relevant patient bag. The feature vector for that read contains one number: the cluster number that the read belongs to.  We compared the runtime and accuracy of the two implementations, which both used a linear kernel. In both cases, the training and test set each contained 30 patients, and each contained 15 patients with type 2 diabetes and 15 control (healthy) patients. The results are discussed in the "Experimental Results" section.

### 2016-08-03 latex/input/intro.tex
added script for generating plots of GICF runs; updated paper

The human body is one of the 
most densely populated microbial 
environments in the world, with human host cells interacting 
with  more than $10^{14}$ microbial cells. Collectively, the human 
and microbial cells are referred to as the 
\emph{human microbiome} \cite{turnbaugh_human_2007,backhed_host-bacterial_2005}. 
%
%
%
Biotechnological  advances 
can 
interrogate the microbial 
communities present across three primary dimensions: (i) clinically different
patients like those suffering from liver disease versus healthy controls (ii) different
patient sites and tissues like skin and the gastrointestinal tract 
and (iii) across different patient visits~\cite{Costello12182009,Qin:2010fk} and before/after treatment. 
%
Methods to query this  microbial
data include DNA sequencing of 
specific marker genes such as the 16S rRNA marker genes~\cite{Woese97}, 
sequencing
the entire pool of microbial genomes at once (metagenomics), or 
sequencing the mixture of 
expressed gene transcripts (metatranscriptomes).
%
These sequencing methods along 
with proteomics and mass spectrometry
allow us to study the
biotransformations caused by the microbial communities. 
%
Several researchers and clinicians have embarked on
studies of the pathogenic and clinical 
role played by the microbiome with respect to
human health and disease conditions.

In this study, we specifically developed
a computational method to predict the clinical phenotype
of a patient i.e., a pipeline to 
predict whether or not the patient has a specific type and level of  a disease.  This 
pipeline
is able to make the phenotypic
predictions from input metagenomic sequences using post-processing 
with unsupervised binning approaches (clustering) or  supervised search based 
taxa profiling approaches.  This method is designed to 
run quickly and produce accurate and sensitive 
results. We  compare and evaluate two state-of-the-art 
approaches, one for binning (UCLUST \cite{Edgar10}) and one for taxonomic profiling (KRAKEN \cite{Wood14}) 
to represent the input metagenome sequences as features for classification. The classification
formulation uses a binary support vector machine based classifier \cite{vap95}. We also 
evaluate the performance of preliminary 
feature scaling and engineering  for classification purposes.

We specifically evaluated our approach on 
microbiome samples of patients suffering from hepatic encephalopathy due 
to liver cirrhosis. 
%
We found that 
UCLUST performed better than Kraken on our data 
set, although Kraken was faster. We were also able to 
engineer the data to improve the accuracy of the classifier. Our 
pipeline had 85.64\% accuracy using UCLUST and 80.66\% accuracy 
using Kraken.

%For our data set with 1.3 million reads, UCLUST took 
%180 seconds to run and Kraken took 120 seconds. 
%

>>>>

The human body contains one of the most dense and diverse microbial environments in the world. The human and microbial cells are collectively referred to as the \emph{human microbiome} \cite{turnbaugh_human_2007,backhed_host-bacterial_2005}. Advances in biotechnology have allowed scientists to directly interrogate the human microbiome, particularly the development of high throughput sequencing technologies, which generate massive amounts of biological data. In recent years, improving capabilities in data science have allowed for the study of \emph{metagenomics}, which involves sequencing the entire pool of microbial genomes at once. This data is usually gathered via \emph{shotgun sequencing}, which generates a number of short \emph{reads} containing bits of genomic data from the microbes of the host environment \cite{messing81}. These reads, represented as strings of nucleotides, represent only small parts of the microbe's full genome, and are not ordered in any way, which presents several challenges that will be discussed further in the paper. Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (phylogeny), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes \cite{handelsman04}. Thus, studying microbial metagenomics is an effective way to predict and model human disease, also known as clinical \emph{phenotype}.

In this study, we develop an efficient classifier that predicts whether or not a patient has a disease based on their microbiome. This can be viewed as a Multiple Instance Learning (MIL) problem, in which we have several \emph{bags} of instances, and we have labels for the bags, but not for each instance within them. In this case, we have a patient (bag) and a label for each patient (whether or not they have a disease), but no labels for each patient's sequence reads (instances). Specifically, we use a Distance-based Bag of Words (D-BoW) method, discussed further in the Background section, which has been shown to be among the most effective and efficient MIL methods \cite{amores13}. MIL has been studied in many contexts, but it has rarely if ever been studied in the context of predicting clinical phenotype based on metagenomic data from the microbiome. However, since datasets in this domains frequently have patient-level labels but almost never have instance-level labels, this is a well-suited domain for multiple instance learning.

We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype. This data is very large (hundreds of gigabytes) and high-dimensional (thousands of dimensions). Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled. \emph{Assembly} is the process of finding pairs of reads in which the end of one read overlaps with the beginning of another, signifying that they are probably contiguous reads from the same genome, and combining them. This process is repeated as much as possible to form long strings called \emph{contigs}. Assembly also reduces the size and dimensionality of the data by discarding reads that cannot be assembled successfully. Most of the discarded reads would not be particularly useful anyway, since the assembly process could not establish their relationship to other reads. The reduction in data size also allows for clustering, which is not feasible for massive datasets. Clustering uses string similarity measures to group similar reads into "clusters", which is a way of identifying which species of microbe each read corresponds to. The alternative, "aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced. From the clustering output, we extract feature vectors, which are then fed into a standard Support Vector Machine (SVM) classifier.

Thus, the entire pipeline consists of assembling the reads of each patient, combining the resulting contigs from each patient into a single file, clustering the contigs, extracting features from the clustering output, and performing classification with the SVM. This process is explained in further detail in the methods section. We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data. We show that our classifier shows significantly improved performance. This is discusses further in the Experimental Results section.

The rest of the paper is organized as follows. Section 2 presents relevant Background information on  Multiple Instance Learning. Section 3 presents the methods we used in the creation of our pipeline. Section 4 presents Materials, such as dataset, hardware, and software descriptions. Section 5 presents our results, and Section 6 presents our conclusions.
